---
title: "Practical Machine Learning Course Project"
author: "Aakanksha Jha"
date: "January 26, 2018"
output: html_document
---

## Introduction & Overview
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. A training set of data of 6 people were given and the goal of this project is to predict (20 cases of testing set) the class of the exercise using the data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. <br>

Data Source for this project <http://groupware.les.inf.puc-rio.br/har> <br>

Six participants performed bicep excercises in five different ways: <br>
1. Class A - Exactly according to the correct specification <br>
2. Class B - Throwing the elbows to the front <br>
3. Class C - Lifting the dumbbell only halfway  <br>
4. Class D - Lowering the dumbbell only halfway <br>
5. Class E - Throwing the hips to the front <br>

# Loading Libraries
```{r, echo=TRUE}
library(knitr)
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
library(corrplot)
```

# Loading Data
```{r, echo=TRUE}
# set the URL for the download

UrlTrain <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
UrlTest  <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

# download the datasets
train <- read.csv(url(UrlTrain))
final_test <- read.csv(url(UrlTest))
```

# Data Exploration
```{r, echo=TRUE}
dim(train)
dim(final_test)
differing <- match(FALSE,(names(final_test) == names(train)))
names(train)[differing]
names(final_test)[differing]
plot(train$classe, xlab="Activity class", ylab="count", main="Distribution of Exercise Method")
```

# Cleaning NAs
```{r, echo=TRUE}
# Remove first five variables which dont contribute for prediction
train <- train[, -(1:5)]
final_test <- final_test[, -(1:5)]

# Removing NAs
NAs <- sapply(train, function(x) mean(is.na(x)))>.95
train<-train[, NAs == FALSE]
final_test<-final_test[, NAs == FALSE]

# Removing Near Zero Variance Variables
nearZeroVariance <- nearZeroVar(train)
train <- train[,-nearZeroVariance]
final_test <- final_test[,-nearZeroVariance]
```

# Creating Training and Testing Set for first level validation
```{r, echo = TRUE}
# Dividing data into 80% training data and 20% testing data
index<- createDataPartition(train$classe, p=4/5, list=FALSE)
training<- train[index, ]
testing <- train[-index, ] # testing set for initial validation and accuracy check purposes

# Checking Dimensions
dim(training) 
dim(testing)
```

# Training Random Forests model
```{r, echo = TRUE}
randomforestmodel <- train(classe ~., method="rf", data=training, trControl=trainControl(method='cv'), number=3, allowParallel=TRUE, importance=TRUE )
randomforestmodel
```

# Validation, Cross Matrix and Accuracy
```{r, echo = TRUE}
validation_initial <- predict(randomforestmodel, testing)
confusion_matrix <- confusionMatrix(testing$classe, validation_initial)
confusion_matrix$table

accuracy <- postResample(testing$classe, validation_initial)
accuracy <- accuracy[[1]]
accuracy <- accuracy*100
paste0("The accuracy of the predicted model is ", accuracy, "%.")
```
The prediction seems to be accurate as we can see in confusion matrix. 

# Application of this model for the given test case
```{r, echo = TRUE}
final_prediction <- predict(randomforestmodel, final_test)
final_prediction
```